M.Sc. Proposal: \\ \bigskip Computational Electrokinetics

### 2012-11-17 notes/thesis-latex-template/thesis.tex
Work on thesis.

\section{The Physical Problem}
Electrokinetic theory describes the dynamics of charged particles
in ionic fluids \cite{masliyah2005book,kirby2010book}.
When a particle acquires surface charge, a layer
of ions of opposite charge is attracted to the surface via    
electric forces, creating a double-layer structure around the
particle (see Figure \ref{fig:EDL}). This structure, called
``Debye layer'', electrically screens the surface charge,
creating a potential difference between the particle and the outer
layer of the fluid bulk.
In cases where the layer width is much smaller than the particle
size, an analytical asymptotic solution for the Debye layer
dynamics can be derived \cite{yariv2010asymptotic}.
\begin{figure}
    \begin{center}
        \includegraphics[width=0.4\textwidth]
            {figs/debye.eps}
        \caption[Schematic structure of the double layer]{
        Schematic structure of the double layer:
        (1) charged particle surface, (2) Debye layer, (3) fluid bulk.
        If the particle surface has positive (black) surface charge,
        it attracts negative (white) ions from the fluid making the
        Debye layer negatively charged (as opposed to the rest of
        the fluid bulk, which is electrically neutral).
        The zeta potential is defined as the voltage drop across 
        the Debye layer (2).}
        \label{fig:EDL}
    \end{center}
\end{figure}
% \begin{tikzpicture}\end{tikzpicture}

The variables of the electrokinetic flow equations are the electrostatic
potential $\varphi$, fluid velocity $\bv$ and its pressure $p$, and
ionic concentration $c$.
The boundary conditions are determined by the specific
problem under consideration and defined by the particle's
geometry, chemical characteristics, and the fluid dynamics.
The partial differential equations that describe the system dynamics
under an external electric field, are coupled and nonlinear, and
in general they have no analytic solution. 
Moreover, any numerical solver must handle the scale disparity caused by the
Debye layer width being much smaller than the particle size. 
An effective boundary condition outside the Debye layer
has been developed for spherical particles, yielding a macroscale 
model at \cite{yariv2010migration, schnitzer2012surface}.
The model is solved for weak electric field in an axisymmetric setting 
but it is hard to extend this analytic solution to more general systems.
Once the electric field becomes stronger, 
significant nonlinear phenomena emerge.
This interesting regime has not yet been explored extensively.
Recently, it has been conceived that such type of dynamics
can be very useful in transporting and manipulating micro-
and nanoscale objects in many nanotechnology applications,
nano-fluidic devices, packed bed separation, desalination processes 
and various electrophoresis applications \cite{howse2007self,paxton2004catalytic,pumera2010electrochemically,nischang2007,leinweber2004}.
Because the nonlinear macroscale model for these electrokinetic
phenomena has no closed-form analytic solution, a numerical
solver for this system is of importance.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
\chapter{Numerical Scheme} \label{ch:algorithm}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Solver Design}
An operator $\cO$ is defined by its input vector $\bx$ and its output 
$\by = \cO(\bx)$.
In order to solve the equation $\cO(\bx) = \bzero$, Newton's method is applied, 
given an initial solution $\bx_0$.
Each step requires the computation of the residual vector, $\br = \cO(\bx)$, and
the gradient matrix of the operator, $\bG = \bnabla \cO(\bx)$, 
given the current solution $\bx$.
Then, a sparse linear system $\bG \Delta \bx = \br$
needs to be solved to yield the update $\Delta \bx$.
When the method converges, $\bx_n \rightarrow \bx_\infty$, 
the asymptotic convergence is in general quadratic:
\begin{equation}
\|\bx_{n+1} - \bx_\infty\| \le k \|\bx_{n} - \bx_\infty\|^2.
\end{equation}
Thus, when the initial guess is close enough to the solution, the solver
will typically converge in very few steps.

@@

\subsubsection{Binary operators}
Let $\cO_1$ and $\cO_2$ be two operators, so their pointwise addition, subtraction,
and multiplication, are defined as follows:
\begin{equation}\begin{array}{cc}
(\cO_1 \pm \cO_2)(\bx) = \cO_1(\bx) \pm \cO_2(\bx), &
\bnabla (\cO_1 \pm \cO_2)(\bx) = \bnabla\cO_1(\bx) \pm \bnabla\cO_2(\bx),
\end{array}\end{equation}
\begin{equation}\begin{array}{cc}
(\cO_1 \cdot \cO_2)(\bx) = \cO_1(\bx) \cdot \cO_2(\bx), &
\bnabla (\cO_1 \cdot \cO_2)(\bx) = \diag(\cO_2(\bx)) \bnabla\cO_1(\bx) 
                                  + \diag(\cO_1(\bx)) \bnabla\cO_2(\bx).
\end{array}\end{equation}
\subsubsection{N-ary operators}
Given $N$ operators $\{\cO_i\}_{i=1}^N$, their concatenation $\cO$ is defined as:
\begin{equation}\begin{array}{cc}
\cO(\bx) = [\cO_1(\bx); \ldots; \cO_N(\bx)], &
(\bnabla\cO)(\bx) = [(\bnabla\cO_1)(\bx); \ldots; (\bnabla\cO_N)(\bx)].
\end{array}\end{equation}

>>>>

\section{The Physical Problem}
\subsection{Electrokinetic Phenomena}
Electrokinetic theory describes the dynamics of charged particles
in ionic fluids \cite{masliyah2005book,kirby2010book}.
When a particle acquires surface charge, a layer
of ions of opposite charge is attracted to the surface via    
electric forces, creating a double-layer structure around the
particle (see Figure \ref{fig:EDL}). This structure, called
``Debye layer'', electrically screens the surface charge,
creating a potential difference between the particle and the outer
layer of the fluid bulk.
In cases where the layer width is much smaller than the particle
size, an analytical asymptotic solution for the Debye layer
dynamics can be derived \cite{yariv2010asymptotic}.
\begin{figure}
    \begin{center}
        \includegraphics[width=0.4\textwidth]
            {figs/debye.eps}
        \caption[Schematic structure of the double layer]{
        Schematic structure of the double layer:
        (1) charged particle surface, (2) Debye layer, (3) fluid bulk.
        If the particle surface has positive (black) surface charge,
        it attracts negative (white) ions from the fluid making the
        Debye layer negatively charged (as opposed to the rest of
        the fluid bulk, which is electrically neutral).
        The zeta potential is defined as the voltage drop across 
        the Debye layer (2).}
        \label{fig:EDL}
    \end{center}
\end{figure}
% \begin{tikzpicture}\end{tikzpicture}

The variables of the electrokinetic flow equations are the electrostatic
potential $\varphi$, fluid velocity $\bv$ and its pressure $p$, and
ionic concentration $c$.
The boundary conditions are determined by the specific
problem under consideration and defined by the particle's
geometry, chemical characteristics, and the fluid dynamics.
The partial differential equations that describe the system dynamics
under an external electric field, are coupled and nonlinear, and
in general they have no analytic solution. 
Moreover, any numerical solver must handle the scale disparity caused by the
Debye layer width being much smaller than the particle size. 
An effective boundary condition outside the Debye layer
has been developed for spherical particles, yielding a macroscale 
model at \cite{yariv2010migration, schnitzer2012surface}.
The model is solved for weak electric field in an axisymmetric setting 
but it is hard to extend this analytic solution to more general systems.
Once the electric field becomes stronger, 
significant nonlinear phenomena emerge.
This interesting regime has not yet been explored extensively.

\subsection{Applications}

Recently, it has been conceived that such type of phenomena
can be very useful in microfluidics
devices and methods for controlling and manipulating fluid flows in micro- and nanoscale. 
Many different components are used in microfluidics such as valves, pumps, sensors, mixers, 
filters, separators, heaters etc., and are combined into lab-on-a-chip systems.

The manipulation can be achieved by either applying the external field at inlets and outlets, or it can be applied locally in the microchannel by integrated components.
A typical microchannel is produced with lithographic techniques. Normally the Reynolds
number is very low (on the order of unity or smaller) which means that viscosity dominates
and the convective term of the Navier–Stokes equations are insignificant. This means that
microflows most often are laminar.

In the last decade there has been an increase in microfluidics research, 
due to the spread of tools for microfluidic systems fabrication, 
production of cheap and portable devices for fast point-of-care analysis, 
and fundamental research of physical, chemical and biological processes.
Various applications are \cite{erickson2004integrated, squires2005microfluidics, hardt2007microfluidic}:
\begin{enumerate}
\item DNA and cell analysis \cite{horsman2007forensic, wu2004chemical}.
\item Separation-based detection \cite{weinberger1993practical}.
\item Microscale chemical processing \cite{kothare2006dynamics}.
\item Packed bed separation \cite{leinweber2004, losey2001microfabricated}.
\item Desalination processes \cite{kim2010direct}.
\item Drug delivery \cite{santini2000microchips}.
\end{enumerate}

Because the nonlinear macroscale model for these electrokinetic
phenomena has no closed-form analytic solution, a numerical
solver for this system is of importance.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Numerical Scheme} \label{ch:algorithm}

@@

We start by presenting the numerical framework used for the nonlinear solver construction.

\subsection{Finite Volume Method}

Finite Volume Methods \cite{ferziger1999computational}
 are based on a discretization of the integral forms of the conservation equations.
The simulation domain is divided into a set of discrete non-overlapping control volumes.
Each conservartion equation is discretized over the control volumes,
such that the integral form of the conservation law is satisfied using an approximation of the fluxes
over the control volume faces.
This method ensures that the discretization is conservative in the discrete sense.

\subsection{Ghost Points}

\subsection{Upwind Discretization}

\subsection{Newton's Method}
Newton's method is used to solve a nonlinear system of equations of the form $\cO(\bx) = \bzero$.
For small step $\Delta\bx$, we can linearize the operator by
\begin{align}
\cO(\bx + \Delta\bx) \approx \cO(\bx) + \bnabla \cO(\bx) \Delta\bx
\end{align}
Thus, we use the following iteration, starting from an initial solution $\bx_0$:
\begin{align}
\Delta\bx_n &= - \left[\bnabla \cO(\bx_n)\right]^{-1} \cdot \cO(\bx_n) \\
\bx_{n+1} &= \bx_n + \Delta\bx_n
\end{align}

Then, a sparse linear system $\boldsymbol{A} \Delta \bx = \boldsymbol{b}$
needs to be solved to yield the update $\Delta \bx$.
When the method converges, $\bx_n \rightarrow \bx_\infty$, 
the asymptotic convergence is in general quadratic:
\begin{equation}
\|\bx_{n+1} - \bx_\infty\| \le k \|\bx_{n} - \bx_\infty\|^2.
\end{equation}
Thus, when the initial guess is close enough to the solution, the solver
will typically converge in very few steps.

\subsection{Richardson Extrapolation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Solver Design}
An operator $\cO$ is defined by its input vector $\bx$ and its output 
$\by = \cO(\bx)$.
In order to solve the equation $\cO(\bx) = \bzero$, Newton's method is applied, 
given an initial solution $\bx_0$.
Each step requires the computation of the residual vector, $\br = \cO(\bx)$, and
the gradient matrix of the operator, $\bG = \bnabla \cO(\bx)$, 
given the current solution $\bx$.

@@

\subsubsection{Binary operators}
Let $\cO_1$ and $\cO_2$ be two operators, so their pointwise addition, subtraction,
and multiplication, are defined as follows:
\begin{align}
(\cO_1 \pm \cO_2)(\bx) &= \cO_1(\bx) \pm \cO_2(\bx), \\
\nonumber
\bnabla (\cO_1 \pm \cO_2)(\bx) &= \bnabla\cO_1(\bx) \pm \bnabla\cO_2(\bx), \\
(\cO_1 \cdot \cO_2)(\bx) &= \cO_1(\bx) \cdot \cO_2(\bx), \\
\nonumber
\bnabla (\cO_1 \cdot \cO_2)(\bx) &= \diag(\cO_2(\bx)) \bnabla\cO_1(\bx) 
                                  + \diag(\cO_1(\bx)) \bnabla\cO_2(\bx).
\end{align}
\subsubsection{N-ary operators}
Given $N$ operators $\{\cO_i\}_{i=1}^N$, their concatenation $\cO$ is defined as:
\begin{align}
\cO(\bx) &= [\cO_1(\bx); \ldots; \cO_N(\bx)], \\
\nonumber
(\bnabla\cO)(\bx) &= [(\bnabla\cO_1)(\bx); \ldots; (\bnabla\cO_N)(\bx)].
\end{align}
